---
title: "RORaFoxp3-microarray-analysis"
author: "artemii"
date: '2022-07-05'
output: html_document
---

```{r setup, include=FALSE}
ROOT_DIR="/Volumes/artemii-4tb/data/microarray/treg-microarray/"
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = ROOT_DIR)
```

## Loading libraries

```{r}
library(limma)
library(affy)
library(statmod)
library(oligo)
library(mogene20sttranscriptcluster.db)
library(genefilter)
```

## Reading data

### Get info about samples

```{r}
targets <- readTargets("config/targets.txt")
targets
```

### Construct expression set: RMA normalize and summarize

Core genes level summarization using median-polish

```{r}
RAW_DATA_PATHS=paste0("raw-data/", targets$FileName)
# Affy commands don't work here because of MoGene 2.0st array, have to use oligo
# eset <- justRMA(filenames = RAW_DATA_PATHS)


rawData <- read.celfiles(RAW_DATA_PATHS)
eset <- oligo::rma(rawData, target = "core")
colnames(eset) <- row.names(targets)
head(exprs(eset))
```

```{r}
dim(eset)
```

### MDS plot

In vitro, eWAT, spleen and lung samples are distinduishable in full MDS plot

```{r}
plotMDS(eset)
```

## Annotation

Select probes with largest variance of those, which map to a single gene

```{r}
stat <- apply(exprs(eset), 1, var)
single_gene_probes <- findLargest(row.names(eset), stat, data = "mogene20sttranscriptcluster")
eset <- eset[row.names(eset) %in% single_gene_probes,]
dim(eset)
```
### ! TODO Average Expression Filtering needed?
```{r}
Anno <- select(mogene20sttranscriptcluster.db, keys = row.names(eset), columns = c("ENTREZID", "SYMBOL", "GENENAME"), keytype = "PROBEID")
Anno <- Anno[!is.na(Anno$ENTREZID),]
keep_probes <- Anno$PROBEID
eset <- eset[row.names(eset) %in% keep_probes, ]
fData(eset) <- Anno[Anno$PROBEID %in% row.names(eset),]
head(fData(eset))
```

```{r}
dim(eset)
```

```{r}
dim(fData(eset))
```

```{r}
pData(eset) <- targets
pData(eset)
```

```{r}
eset
```

## Multi-factor mouse blocking linear model 
### Prepare design matrix

```{r}
Mouse <- as.factor(targets$Mouse)

Genotype <- as.factor(targets$Genotype)
Genotype <- relevel(Genotype, ref = "WT")
Tissue <- as.factor(targets$Tissue)
Tissue <- relevel(Tissue, ref = "Spleen")
Experiment <- as.factor(targets$Experiment)
Treatment <- as.factor(targets$Treatment)
Group <- as.factor(targets$Group)

Condition <- factor(paste(targets$Genotype, targets$Tissue,
                          targets$Treatment, sep = "."),
                    levels = c("WT.Spleen.CD", "KO.Spleen.CD",
                               "WT.eWAT.HFD", "KO.eWAT.HFD",
                               "WT.Lung.HDM", "KO.Lung.HDM",
                               "WT.mLN.HDM", "KO.mLN.HDM",
                               "WT.in_vitro.IL2", "KO.in_vitro.IL2",
                               "WT.in_vitro.IL2_IL4", "KO.in_vitro.IL2_IL4")
                    )

design <- model.matrix(~ 0 + Condition)
colnames(design) <- levels(Condition)
head(design)
```

```{r}
dim(design)
```

```{r}
colnames(design)
```

Just testing if there is no confused samples

```{r}
test_filenames <- targets$FileName[as.logical(design[,"WT.Lung.HDM"])]
targets[targets$FileName %in% test_filenames,]
```


```{r}
contrast.matrix <- makeContrasts(
  KOvsWTin_vitro = (KO.in_vitro.IL2_IL4-KO.in_vitro.IL2) - 
    (WT.in_vitro.IL2_IL4-WT.in_vitro.IL2),
  levels = design
)
```



### Compute correlations for measurements made on the same mice
```{r}
corfit <- duplicateCorrelation(eset, design, block = targets$Mouse)
corfit$consensus
```


### Fit a linear model with contrasts and blocking

```{r}

fit <- lmFit(eset, design, block = targets$Mouse, correlation = corfit$consensus)
multi_mouse <- fit
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)
topTable(fit2, coef = "KOvsWTin_vitro")
```
## Multi-factor experiment blocking linear model
### Prepare design matrix

```{r}
Experiment <- as.factor(targets$Experiment)
Condition <- factor(paste(targets$Genotype, targets$Tissue,
                          targets$Treatment, sep = "."),
                    levels = c("WT.Spleen.CD", "KO.Spleen.CD",
                               "WT.eWAT.HFD", "KO.eWAT.HFD",
                               "WT.Lung.HDM", "KO.Lung.HDM",
                               "WT.mLN.HDM", "KO.mLN.HDM",
                               "WT.in_vitro.IL2", "KO.in_vitro.IL2",
                               "WT.in_vitro.IL2_IL4", "KO.in_vitro.IL2_IL4")
                    )
design <- model.matrix(~ 0 + Condition)
colnames(design) <- levels(Condition)
colnames(design)
```

### Compute correlations for technical replicates
```{r}
corfit <- duplicateCorrelation(eset, design, block = targets$Experiment)
corfit$consensus
```

### Fit a linear model

```{r}
fit <- lmFit(eset, design, block = targets$Experiment, correlation = corfit$consensus.correlation)
multi_exp <- fit
contrast.matrix <- makeContrasts(
  KOvsWTinIL2 = KO.in_vitro.IL2-WT.in_vitro.IL2,
  KOvsWTinIL2_IL4 = KO.in_vitro.IL2_IL4-WT.in_vitro.IL2_IL4,
  Diff_IL4vsIL2 = (KO.in_vitro.IL2_IL4-KO.in_vitro.IL2) - (WT.in_vitro.IL2_IL4-WT.in_vitro.IL2),
  KOvsWTineWAT = KO.eWAT.HFD-WT.eWAT.HFD,
  KOvsWTinLung = KO.Lung.HDM-WT.Lung.HDM,
  KOvsWTinmLN = KO.mLN.HDM-WT.mLN.HDM,
  KOvsWTinSpleen = KO.Spleen.CD-WT.Spleen.CD,
  Diff_mLNvsSpleen = (KO.mLN.HDM-KO.Spleen.CD) - (WT.mLN.HDM-WT.Spleen.CD),
  levels = design
)
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)
```

#### IL2 in vitro samples
```{r}
topTable(fit2, coef = "KOvsWTinIL2", number = 500)
```

#### IL2 + IL4 in vitro samples
```{r}
topTable(fit2, coef = "KOvsWTinIL2_IL4", number = 500)
```

#### Response to IL4 in KOvsWT
```{r}
topTable(fit2, coef = "Diff_IL4vsIL2", number = 500)
```
#### Epididymal White Adipose Tissue
```{r}
topTable(fit2, coef = "KOvsWTineWAT", number = 500)
```

#### Lung
```{r}
topTable(fit2, coef = "KOvsWTinLung", number = 500)
```

#### Mesenteric lymph nodes
```{r}
topTable(fit2, coef = "KOvsWTinmLN", number = 500)
```

#### Spleen
```{r}
topTable(fit2, coef = "KOvsWTinSpleen", number = 500)
```

#### Activated vs normal lymphoid Treg
```{r}
topTable(fit2, coef = "Diff_mLNvsSpleen", number = 500)
```

## Nested interaction linear model
At first I will try a model with no blocking
### Prepare design matrix
Here each condition is compared to CD Spleen Treg
I'm not sure it's the best option
```{r}
Condition <- factor(paste(targets$Tissue,targets$Treatment, sep = "."),
                    levels = c("Spleen.CD",
                               "eWAT.HFD",
                               "Lung.HDM",
                               "mLN.HDM",
                               "in_vitro.IL2",
                               "in_vitro.IL2_IL4")
                    )
Genotype <- as.factor(targets$Genotype)
Genotype <- relevel(Genotype, ref = "WT")

design <- model.matrix(~Genotype + Genotype:Condition)
colnames(design)
```

### Fit a linear model

#### IL2 in vitro samples
```{r}
fit <- lmFit(eset, design)
nested_fit <- fit
fit <- eBayes(fit)
fit2 <- contrasts.fit(fit, c(0, 0, 0, 0, 0, 0,
                             0, 0,-1, 1, 0, 0))
fit2 <- eBayes(fit2)
topTable(fit2, number = 100, adjust.method = "BH")
```
#### IL2 + IL4 in vitro samples
```{r}
fit <- lmFit(eset, design)
fit <- eBayes(fit)
fit2 <- contrasts.fit(fit, c(0,0,0,0,0,0,
                             0,0,0,0,-1,1))
fit2 <- eBayes(fit2)
topTable(fit2, number = 100)
```
#### Epididymal White Adipose Tissue
```{r}
fit2 <- contrasts.fit(fit, c(0, 0,-1, 1, 0, 0,
                             0, 0, 0, 0, 0, 0))
fit2 <- eBayes(fit2)
topTable(fit2, number = 100)
```

#### Lung
```{r}
fit2 <- contrasts.fit(fit, c(0, 0, 0, 0,-1, 1,
                             0, 0, 0, 0, 0, 0))
fit2 <- eBayes(fit2)
topTable(fit2, number = 100)
```
#### Mesenteric lymph nodes
```{r}
fit2 <- contrasts.fit(fit, c(0, 0, 0, 0, 0, 0,
                            -1, 1, 0, 0, 0, 0))
fit2 <- eBayes(fit2)
topTable(fit2, number = 100)
```


#### Spleen
```{r}
fit <- eBayes(fit)
topTable(fit, coef = 2, number = 100)
```

## Single-factor with blocking
This one doesn't work
### ! TODO Why doesn't work??
For in vitro samples there are three experiments. This might create batch effect. I'll try to include it in the linear model.
### ! TODO Is it okay to do blocking with all the conditions?
### Prepare design matrix
```{r}

Experiment <- as.factor(targets$Experiment)
c <- factor(paste(targets$Genotype, targets$Tissue,
                  targets$Treatment, sep = "."),
            levels = c("WT.Spleen.CD", "KO.Spleen.CD",
                       "WT.eWAT.HFD", "KO.eWAT.HFD",
                       "WT.Lung.HDM", "KO.Lung.HDM",
                       "WT.mLN.HDM", "KO.mLN.HDM",
                       "WT.in_vitro.IL2", "KO.in_vitro.IL2",
                       "WT.in_vitro.IL2_IL4", "KO.in_vitro.IL2_IL4")
                    )

design <- model.matrix(~Experiment + c)
colnames(design)
```

### Fit a linear model
```{r}
fit <- lmFit(eset, design)
```



