---
title: "RORaFoxp3-microarray-analysis"
author: "artemii"
date: '2022-07-05'
output: html_document
---


```{r setup, include=FALSE}
ROOT_DIR="/Volumes/artemii-4tb/data/rora-foxp3-genomics/microarray/treg-microarray"
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = ROOT_DIR)
```

## Loading libraries

```{r}
library(limma)
library(affy)
library(statmod)
library(oligo)
library(mogene20sttranscriptcluster.db)
library(genefilter)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(svglite)

```

## Reading data

### Get info about samples

```{r}
targets <- readTargets("config/targets.txt")
targets
```

### Construct expression set: RMA normalize and summarize

Core genes level summarization using median-polish

```{r}
RAW_DATA_PATHS=paste0("raw-data/", targets$FileName)
# Affy commands don't work here because of MoGene 2.0st array, have to use oligo
# eset <- justRMA(filenames = RAW_DATA_PATHS)


rawData <- read.celfiles(RAW_DATA_PATHS)
eset <- oligo::rma(rawData, target = "core")
colnames(eset) <- row.names(targets)
head(exprs(eset))
```

## QC
### MA plots per tissue
```{r}

invitro_indeces <- 1:20
ewat_indeces <- 21:26
spleen_indeces <- 27:34
lung_indeces <- 35:40

tissue_index_list <- list(invitro_indeces,
                          ewat_indeces,
                          spleen_indeces,
                          lung_indeces)

for (tissue in tissue_index_list){
  getwd()
  eset_tissue_slice <- eset[, tissue]
  
  par(mar = c(1, 1, 1, 1))
  xl <- c(2.8, 4)
  yl <- c(-1, 1)
  MAplot(eset_tissue_slice,
         pairs = TRUE,
         ylim = yl,
         xlim = xl,
         plotFun = plot)
  }

```

### Box plots

```{r}
oligo::boxplot(eset)
```

### Density plots
```{r}
oligo::hist(eset, transfo=log2)
```

### Array quality

```{r}
arrayw <- arrayWeights(eset)
barplot(arrayw, xlab="Array", ylab="Weight", col="white", las=2)
abline(h=1, lwd=1, lty=2)
```

### MDS plot

In vitro, eWAT, spleen and lung samples are distinduishable in full MDS plot

```{r}
plotMDS(eset)
```

## Annotation

Select probes with largest variance of those, which map to a single gene

```{r}
stat <- apply(exprs(eset), 1, var)
single_gene_probes <- findLargest(row.names(eset), stat, data = "mogene20sttranscriptcluster")
eset <- eset[row.names(eset) %in% single_gene_probes,]
dim(eset)
```

```{r}
Anno <- AnnotationDbi::select(mogene20sttranscriptcluster.db, keys = row.names(eset), columns = c("ENTREZID", "SYMBOL", "GENENAME"), keytype = "PROBEID")
Anno <- Anno[!is.na(Anno$ENTREZID),]
keep_probes <- Anno$PROBEID
eset <- eset[row.names(eset) %in% keep_probes, ]
fData(eset) <- Anno[Anno$PROBEID %in% row.names(eset),]
head(fData(eset))
```

```{r}
dim(eset)
```

```{r}
dim(fData(eset))
```

```{r}
pData(eset) <- targets
pData(eset)
```

```{r}
eset
```

## Multi-factor experiment blocking linear model
### Prepare design matrix

```{r}
Experiment <- as.factor(targets$Experiment)
Condition <- factor(paste(targets$Genotype, targets$Tissue,
                          targets$Treatment, sep = "."),
                    levels = c("WT.Spleen.CD", "KO.Spleen.CD",
                               "WT.eWAT.HFD", "KO.eWAT.HFD",
                               "WT.Lung.HDM", "KO.Lung.HDM",
                               "WT.mLN.HDM", "KO.mLN.HDM",
                               "WT.in_vitro.IL2", "KO.in_vitro.IL2",
                               "WT.in_vitro.IL2_IL4", "KO.in_vitro.IL2_IL4")
                    )
design <- model.matrix(~ 0 + Condition)
colnames(design) <- levels(Condition)
colnames(design)
```

### Compute correlations for technical replicates
```{r}
corfit <- duplicateCorrelation(eset, design, block = targets$Experiment)
corfit$consensus
```

### Fit a linear model

```{r}
fit <- lmFit(eset, design, block = targets$Experiment, correlation = corfit$consensus.correlation)
multi_exp <- fit
contrast.matrix <- makeContrasts(
  KOvsWTinIL2 = KO.in_vitro.IL2-WT.in_vitro.IL2,
  KOvsWTinIL2_IL4 = KO.in_vitro.IL2_IL4-WT.in_vitro.IL2_IL4,
  Diff_IL4vsIL2 = (KO.in_vitro.IL2_IL4-KO.in_vitro.IL2) - (WT.in_vitro.IL2_IL4-WT.in_vitro.IL2),
  KOvsWTineWAT = KO.eWAT.HFD-WT.eWAT.HFD,
  KOvsWTinLung = KO.Lung.HDM-WT.Lung.HDM,
  KOvsWTinmLN = KO.mLN.HDM-WT.mLN.HDM,
  KOvsWTinSpleen = KO.Spleen.CD-WT.Spleen.CD,
  Diff_mLNvsSpleen = (KO.mLN.HDM-KO.Spleen.CD) - (WT.mLN.HDM-WT.Spleen.CD),
  levels = design
)
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)
```


#### IL2 in vitro samples
```{r}
topTable(fit2, coef = "KOvsWTinIL2", number = 500)
```

#### IL2 + IL4 in vitro samples
```{r}
topTable(fit2, coef = "KOvsWTinIL2_IL4", number = 500)
```

#### Response to IL4 in KOvsWT
```{r}
topTable(fit2, coef = "Diff_IL4vsIL2", number = 500)
```
#### Epididymal White Adipose Tissue
```{r}
topTable(fit2, coef = "KOvsWTineWAT", number = 500)
```

#### Lung
```{r}
topTable(fit2, coef = "KOvsWTinLung", number = 500)
```

#### Mesenteric lymph nodes
```{r}
topTable(fit2, coef = "KOvsWTinmLN", number = 500)
```

#### Spleen
```{r}
topTable(fit2, coef = "KOvsWTinSpleen", number = 500)
```

#### Activated vs normal lymphoid Treg
```{r}
topTable(fit2, coef = "Diff_mLNvsSpleen", number = 500)
```


## Visualization
### Volcano plot function

```{r}
plot_volcano <- function(exp_data, top_label = 5,
                         lim_logFC = F,
                         lim_FDR = F,
                         lim_up_x = NULL,
                         lim_dn_x = NULL,
                         logfc_fdr_thresholds_for_labels = NULL,
                         additional_genes_to_label = c()) {

  
  exp_data <- exp_data %>%
    mutate(
      Significance = case_when(
        abs(logFC) >= log(2, base = 2) & P.Value <= 0.05 ~ "FDR_0_05",
        TRUE ~ "UNCHANGED"
      )
    )
  

  if (length(logfc_fdr_thresholds_for_labels) != 0) {
    logfc_threshold = logfc_fdr_thresholds_for_labels[1]
    fdr_threshold = logfc_fdr_thresholds_for_labels[2]
    exp_data <- exp_data %>%
      mutate(
        Expression = case_when(logFC >= logfc_threshold & P.Value <= fdr_threshold ~ "Up-regulated",
                               logFC <= - logfc_threshold & P.Value <= fdr_threshold ~ "Down-regulated",
                               TRUE ~ "Unchanged")
      )
  } else {
    logfc_threshold = log(2, base = 2)
    fdr_threshold = 0.05
    exp_data <- exp_data %>%
      mutate(
        Expression = case_when(logFC >= logfc_threshold & P.Value <= fdr_threshold ~ "Up-regulated",
                             logFC <= -logfc_threshold & P.Value <= fdr_threshold ~ "Down-regulated",
                             TRUE ~ "Unchanged")
      )
    }
    
  top <-  top_label
  top_genes_df <- bind_rows(
    exp_data %>%
      filter(Expression == 'Up-regulated') %>%
      arrange(P.Value, desc(abs(logFC))) %>%
      head(top),
    exp_data %>%
      filter(Expression == 'Down-regulated') %>%
      arrange(P.Value, desc(abs(logFC))) %>%
      head(top),
  )
  
  genes_to_label_df <- bind_rows(
    top_genes_df,
    exp_data %>%
      filter(SYMBOL %in% additional_genes_to_label)
  )
  
  print(head(genes_to_label_df, n = top))
  if (is.null(lim_up_x)){
    lim_up_x <- max(genes_to_label_df$logFC) * 1.2
  }
  
  if (is.null(lim_dn_x)){
    lim_dn_x <- min(genes_to_label_df$logFC) * 1.2
  }
  
  lim_up_y <- -log(min(genes_to_label_df$P.Value), 10) * 1.2
    
  lim_x <- max(c(abs(lim_up_x), abs(lim_dn_x)))
  
  num_up_regulated <- nrow(filter(exp_data, Expression == 'Up-regulated'))
  print(filter(exp_data, Expression == 'Up-regulated'))
  num_dn_regulated <- nrow(filter(exp_data, Expression == 'Down-regulated'))
  print(filter(exp_data, Expression == 'Down-regulated'))
  
   ggplot(exp_data, aes(x = logFC, y = -log(P.Value, 10))) +
     geom_point(aes(color = Expression), size = 2) +
     
     scale_x_continuous(expand = c(0, 0), limits = c(lim_dn_x, lim_up_x)) +
     scale_y_continuous(expand = c(0, 0), limits = c(0, lim_up_y)) +
     
     
     scale_color_manual(values = c("Unchanged" = "#5C88DA",
                                   "Up-regulated" = "#CC0C00",
                                   "Down-regulated" = "#CC0C00")) +
     
     
     geom_vline(xintercept=c(-logfc_threshold, logfc_threshold),
                col="grey",
                linetype = "dashed",
                size = 0.8) +
     geom_hline(yintercept=-log10(fdr_threshold),
                col="grey",
                linetype = "dashed",
                size = 0.8) +
     annotate(geom = "label",
              x = c(- logfc_threshold, logfc_threshold),
              y = lim_up_y - 0.2,
              label = c(- logfc_threshold, logfc_threshold),
              fill = "white",
              color =  "grey",
              size = 5,
              label.size = NA,
              label.padding = unit(0, "lines")) +
     
     
     geom_text_repel(data = genes_to_label_df,
                     aes(logFC, -log(P.Value, 10), label = SYMBOL),
                     size = 6,
                     max.overlaps = 20,
                     min.segment.length = 0,
                     segment.size = 1,
                     #nudge_x = - 0.2,
                     #force_pull = - 0.007
                     ) +
     
     
     annotate(geom = "text",
              x = lim_x - 0.5,
              y = lim_up_y - 0.2,
              label = paste0("Up: ", num_up_regulated),
              size = 5,
              color = "grey") +
     annotate(geom = "text",
              x = lim_dn_x + 0.5,
              y = lim_up_y - 0.2,
              label = paste0("Down: ", num_dn_regulated),
              size = 5,
              color = "grey") +
     annotate(geom = "text",
              x = lim_dn_x + 0.45,
              y = -log10(fdr_threshold) - 0.2,
              label = paste0("P = ", fdr_threshold),
              size = 5,
              color = "grey") +

     theme_classic() +
     theme(legend.position = "none")
   }
```


```{r}
diff_exp_data <- topTable(fit2, coef = "KOvsWTinIL2_IL4", number = 20000)



```

```{r}
diff_exp_data_paper <- read_delim("/Users/artemii/Desktop/exprs_IL4.csv", delim = ";")
```


```{r}
diff_exp_data_old <- na.omit(read_delim("/Volumes/artemii-4tb/data/desktop-archive/desktop_archive_20250128/KO_act_IL4-WT_act_IL4(ComBat)-31-may-2020.txt"))
```

```{r}
diff_exp_data_combat <- na.omit(read_delim("/Volumes/artemii-4tb/data/rora-foxp3-genomics/microarray/old-version-analysis/RORaTregMA/test_20250105.txt"))
```

#### Plot

```{r}
plot_volcano(diff_exp_data_old,
             top_label = 0,
             lim_up_x = 3,
             lim_dn_x = -2,
             logfc_fdr_thresholds_for_labels = c(0.58, 0.05),
             additional_genes_to_label = c("Bach2", "Cd226", "Il5", "Prf1", "Il3",
                                           "Csf2", "Gzme", "Gzmd", "Gzmf", "Gzmg",
                                           "Gstm5", "Trim30d", "Gm23482"
                                           ))
```

```{r}
ggsave("~/Desktop/volcano_IL4_act.svg",
       width = 1572,
       height = 1456,
       units = "px",
       device = svglite,)
```

```{r}
up <- filter(diff_exp_data_old, SYMBOL %in% roraDE.up)
microarray1 <- c("Gm23193", "Gm23540", "Mir103-2", "Gm23482", "Gm13251", "Gm23205", "Cd226", "BE692007", "Muc13", "Mir181b-2", "Gm22005", "Gm22908", "Gm23256", "Gm26140", "Gm25654", "2310040G07Rik", "Gad2", "Gm25321", "Tgtp1", "A630023P12Rik", "Trat1", "Gm23058", "9930111J21Rik2", "Mir3967", "Gm25973", "Gm8369", "Gm25506", "Tlr1", "Parp8", "Gm24431", "Gm24775", "Gm24622", "Npy", "Gm22213", "Csn2", "Gm24927", "Gm23557", "Gm22322", "Slamf6", "Pisd-ps3", "C230066G23Rik", "Sell", "Scarna3b", "Lrrc4", "Gm24569", "2310044K18Rik", "Lncpint", "Atxn7l1os1", "Cxcr1", "Gm22446", "Gm25636", "Scarna3a", "Id3", "Il1rl1", "Mirlet7a-1", "Mir3074-1", "Nt5e", "Tgtp2", "Snord17", "A630089N07Rik", "Tcf7", "Gm22304", "Snord35b", "Frmd5", "Spef2", "Mir17", "Zbtb10", "Trbv23", "Bach2os", "Gm26514", "Tbrg3", "Snord45b", "Gpr83", "Atp10d", "Gm22272", "Gm17218", "Gm25945", "Snord92", "Gm22887", "Gm26419", "Tnfsf4", "Gm25109", "Gm22960", "H2-Aa", "Mir1967", "Zfp874a", "Ikzf2", "Gm22890", "Ercc6l2", "Airn", "Gm26187", "Gm26666", "Slc25a36", "Kcnq1ot1", "Rev3l", "Cpm", "Nrn1", "Zfp831", "Dpp4", "Gm25518", "Gm25515", "Mir17hg", "Gm24142", "2610204G07Rik", "Gdap10", "Gm24878", "Mir1955", "2010016I18Rik", "Tnfrsf26", "Zfp141", "LOC105247533", "Pik3c2a", "Trav6-1", "Gm26255", "Adamtsl4", "Arap2", "Gm26838", "Clec2i", "A430078G23Rik", "Hc", "Gm23513", "Snord37", "Plcxd2", "Gm22305", "Syp", "Olfr99", "Gm26265", "Kdm7a", "Igf1r", "Gm25779", "Adcy4", "Gpnmb", "Gm25913", "Rab6b", "Mir5098", "Arnt2", "Gm25323", "Fam196b", "Gm25196", "Hddc3", "Gm13833", "n-R5s192", "P2rx7", "Gm22572", "Gm16158", "Gm20456", "Plcl1", "2010300F17Rik", "Gm24813", "Gm23054", "Gm24876", "n-R5s33", "Gm22977", "4930595D18Rik", "Ar", "2010111I01Rik", "Smarca5-ps", "n-R5s39", "E030011O05Rik", "St8sia4", "1700054O19Rik", "Itga4", "Neb", "Mir1198", "Gm12000", "Gm26877", "Gm22362", "Gm22200", "1700109H08Rik", "Gimap7", "AI987944", "Gm22925", "Pln", "Gm24445", "Gja1", "Bach2", "Gm16085", "Letm2", "Crebrf", "B230325K18Rik", "Traj61", "Gm20236", "Fbxl4", "Hpgds", "4930403D09Rik", "Smad3", "Fam65b", "Lpin1", "Mtmr7", "Htr2b", "Gm23297", "Tgfbr1")

microarray2 <- c("Ago3", "Gbp4", "Klhl24", "Tanc2", "Gm16523", "Gm15777", "S1pr1", "Zfp329", "4930431P03Rik", "Mplkip", "Simc1", "Pikfyve", "Gm25882", "Gm15573", "Eif3j1", "Fktn", "Rab39b", "Gm23734", "Nf1", "Upf2", "Eml5", "Gm20695", "Gm24053", "Bmp7", "Akap9", "Gm24119", "Gm20139", "Gm23206", "Ankrd26", "Gm3383", "Mreg", "Gm24938", "Zdhhc23", "Mbtd1", "Gm12185", "Gm11672", "Zfp397", "Gm26225", "Trav9-1", "Zfp442", "D630008O14Rik", "Bmpr2", "Ly75", "Klf7", "Snord53", "Map1b", "Trav5-1", "Fnip2", "Gm24411", "Rictor", "2810474O19Rik", "Gm11465", "Rev1", "Gm24916", "Herc1", "Mr1", "A830080D01Rik", "LOC73899", "Gm22738", "Mir18", "n-R5s80", "Stra6", "Lama5", "Macf1", "Plac8", "Entpd7", "Gm22327", "Ptges3l", "Hscb", "Ocel1", "Eci1", "Mir32", "Cyp11a1", "Tmem138", "Gm6594", "Gm25244", "Ube2t", "Gm8973", "Cks2", "Gm12891", "Rpl29-ps5", "Shcbp1", "Mir684-2", "Tsx", "C4a", "Defa25", "Suox", "Ska3", "Prkaca", "Nr4a3", "Gm23500", "S100a1", "Gm22224", "Idi2", "Gm12166", "Gm25865", "Tprkb", "Kifc1", "Adgrg5", "Adck1", "Rgs16", "Igf2bp2", "Gm25990", "mt-Tr", "Sema6d", "Ppp1r2-ps3", "Erdr1", "Emilin2", "Gm8724", "Cdkn3", "Lce1h", "Cdkn2b", "Gm25720", "Cdc20", "Sema7a", "Chst11", "Traj44", "Spats2", "Gm23723", "Galnt3", "Gm6685", "Nudt14", "Olfr820", "Gm14029", "Gm21887", "Clnk", "Hip1r", "Mnd1", "Ndufb3", "Gm19705", "Ighv2-6-8", "Cd8b1", "mt-Tm", "Sgms2", "1700012B09Rik", "Gm25663", "Tspan6", "Traj40", "Gm5327", "Gm25227", "Gm6195", "Dnajb13", "mt-Ty", "mt-Tt", "Gm16494", "Ccl4", "Gm7676", "Gm10320", "Isg15", "Pilrb2", "Lrrk1", "Traj53", "Kcnk7", "Gm22723", "Spata24", "Gm25989", "Lag3", "Ppm1h", "Gm25985", "mt-Tf", "Xdh", "Plk1", "mt-Tc", "Gm23389", "Gm15473", "Gzmk", "Gm25360", "Snora30", "Ccnb2", "Olfr1269", "Batf3", "Vash1", "Fkbp11", "Apoa2", "mt-Tk", "B4galt5", "Gpr141", "Hmga1", "Lax1", "Rab27a", "Gm13275", "Adora2b", "Spry4", "Ifng", "Traj56", "mt-Tq", "Gm7075", "Havcr2", "Cldnd2", "Myl6b", "Sult2b1", "Pgam1", "Apoo-ps", "Ighv14-4", "Rpl7a-ps5", "Gch1", "Pxdc1", "Trav6d-4", "Nkg7", "Spry2", "Dennd3", "Gm23444", "Gm24497", "Gm25939", "Trav13d-4", "Avil", "Trav5-4", "Acbd7", "Pak6", "Pmf1", "Gm22369", "Plek", "Gm24415", "Slamf7", "Zbtb32", "Rpl21-ps7", "Cdc25c", "n-R5s28", "Psmb5-ps", "Tnp2", "Tjp1", "Klrc1", "1500009L16Rik", "Cd40lg", "Zfp683", "Clybl", "Arsb", "Gm6625", "Slc37a2", "Hist1h3e", "Gem", "Gm5566", "Gm3238", "Gm26088", "Nkain1", "Gm15135", "Gm25813", "Gm11937", "Mpzl2", "Rps13-ps1", "S100a5", "Spp1", "Il31", "Gm16224", "mt-Tv", "Trim12a", "AF357399", "Hist1h2ag", "Acpp", "Ly6c2", "Ccl3", "Gm8871", "Ccl5", "Klrd1", "Cd8a", "Pcbd2", "Gpr33", "Khdc1c", "Hid1", "Gm26014", "Hist1h2bk", "Gm10566", "Magohb", "mt-Tg", "Gzmc", "Ifitm1", "Mir1932", "Pdcd1", "Gm8922", "Xcl1", "Gm5152", "Gm4609", "Cenpw", "Gm6460", "Hist1h2bc", "Il1a", "Ly6g5b", "Crabp2", "Gm8906", "Gm6455", "Tmem171", "Insl6", "Fam20a", "Fasl", "Srgap3", "Gm8879", "Il17a", "Il2", "Cd160", "Speer1", "Atp8b4", "Gm2663", "LOC102638993", "Scimp", "Il6", "Gm26527", "Cd244", "Eomes", "Gm12214", "Il23a", "4933402N22Rik", "Crtam", "Il3", "Il5", "Gstm5", "Fcer1g", "Prf1", "Csf2", "Gm1966", "Klra5", "Trim30d", "Akr1c18", "Gzmg", "Hbb-bt", "Gzmf", "Gzmd", "Gzme")

microarray <- c(microarray1, microarray2)


de1 <- filter(diff_exp_data_old, SYMBOL %in% microarray)

```


```{r}

up <- filter(de1, logFC > 0.58)
```

